---
title: "DOM Experiment Analysis Notebook"
author: "Cat Luria"
date: "July 15, 2015"
output: 
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes

---
\newpage
```{r, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```
---


#Getting started

First, load all the necessary packages.

```{r libraries, echo=FALSE, warning=FALSE}
#libraries
library(data.table)
library(BiodiversityR)
library(vegan)
library(ggplot2)
library(wesanderson)
library(scales)
library(reshape2)
library(RColorBrewer)
library(gdata)
library(plyr)
library("phyloseq")
library(scales)
library(grid)
library(Rcpp)
library(RcppEigen)
```

```{r multiplot_function, echo=FALSE}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```


```{r get_data, echo=FALSE}

setwd("/Users/cmluria/Documents/Documents/Thesis Chapters/DOM/16S_library_analysis")

otutable <- fread('cdout_DOM_bac/otu_table_mc2_DOM_even64271_bac_from_biom.txt', skip=1)

meta <- fread('DOM.map.txt')

#order meta according to MBL_library ID
idx_meta <- order(meta$'#SampleID', decreasing=FALSE)
meta_sorted <- meta[idx_meta]

```

```{r remove_chloroplasts, echo=FALSE}
#creating a table that has chloroplasts removed
#first create index of all rows with "organelle"
#then invert, to get index of rows without "organelle"
#apply index to original table
chlor <- grep('Chloroplast', x=otutable$taxonomy)
idx_chlor <- grep('Chloroplast',x = otutable$taxonomy, invert=TRUE)
otutable_bac <- otutable[i=idx_chlor]
```

#Alpha Diversity

```{r manip_otus, echo=FALSE}

#convert the otu table to a data frame
otutable_bac <- as.data.frame(otutable_bac)
#grab the columns with data
numberofsamples <- dim(otutable_bac)[2] - 1
otutable_bac_samples <- otutable_bac[,(2:numberofsamples)]
#transpose
otutable_bac_samples_t <- t(otutable_bac_samples)
#use column 1 from original table as column names in new table
colnames(otutable_bac_samples_t) <- otutable_bac[,1]

new_otutable_bac_samples_t <- otutable_bac_samples_t[order(row.names(otutable_bac_samples_t)), ]
otutable_bac_samples_t <- as.data.frame(new_otutable_bac_samples_t)
```

```{r calculate_richness, echo=FALSE}

# otutable_meta_DOM_df <- as.data.frame(otutable_meta_DOM)
# otutable_samples_DOM <- otutable_meta_DOM_df[,-(18208:18232)]

#calculate richness by site
richness <- diversityresult(otutable_bac_samples_t, 
                              index='richness', method='s')
#add metadata to richness
diversitytable <- cbind(meta_sorted, richness)

```

```{r anova_DOM, echo=FALSE, include=FALSE}
# aov.richness.month= aov(richness~Month,data=diversitytable_SWI)
# summary(aov.richness.month)
# pvalue <- summary(aov.richness.month)[[1]][, 5][1]
# 
# tuk<- TukeyHSD(aov.richness.month)
# tuk

```

```{r aggregate_richness_by_treatment, echo=FALSE, fig.cap="Changes in richness under different experimental treatments."}

cdata <- ddply(diversitytable, c("Experiment", "Timepoint", "Treatment"), summarise,
               N = length(richness),
               mean = mean(richness),
               sd   = sd(richness),
               se   = sd / sqrt(N)
)

cdata <- as.data.table(cdata)
#cdata$Experiment <- as.factor(cdata$Experiment)
cdata1 <- cdata[which(cdata$Experiment=="1")]
cdata2 <- cdata[which(cdata$Experiment=="2")]
cdata3 <- cdata[which(cdata$Experiment=="3")]
cdata4 <- cdata[which(cdata$Experiment=="4")]

p1 <- ggplot(cdata1, aes(x=Timepoint, y=mean, colour=Treatment)) + 
    geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.1) +
    geom_line() +
    geom_point() +
    scale_x_continuous(name="Timepoint") +
    scale_y_continuous(name="Richness") +
    ggtitle("August") +
    theme_bw() 
    #theme(legend.position="none")
   
p2 <- ggplot(cdata2, aes(x=Timepoint, y=mean, colour=Treatment)) + 
    geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.1) +
    geom_line() +
    geom_point() +
    scale_x_continuous(name="Timepoint") +
    scale_y_continuous(name="Richness") +
    ggtitle("September") +
    theme_bw() 
    #theme(legend.position="none")
   
p3 <- ggplot(cdata3, aes(x=Timepoint, y=mean, colour=Treatment)) + 
    geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.1) +
    geom_line() +
    geom_point() +
    scale_x_continuous(name="Timepoint") +
    scale_y_continuous(name="Richness") +
    ggtitle("October") +
    theme_bw() 
    #theme(legend.position="none")

p4 <- ggplot(cdata4, aes(x=Timepoint, y=mean, colour=Treatment)) + 
    geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.1) +
    geom_line() +
    geom_point() +
    scale_x_continuous(name="Timepoint") +
    scale_y_continuous(name="Richness") +
    ggtitle("December") +
    theme_bw() 
    #theme(legend.position="none")
   
multiplot(p1, p2, p3, p4, cols=2)

```

\newpage

```{r aug_exp1_richness_plot}
ggplot(cdata1, aes(x=Timepoint, y=mean, colour=Treatment)) + 
    geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.1) +
    geom_line() +
    geom_point() +
    scale_x_continuous(name="Timepoint") +
    scale_y_continuous(name="Richness") +
    theme_bw() 
    #theme(legend.position="none")
```

```{r dec_exp4_richness_plot}
ggplot(cdata4, aes(x=Timepoint, y=mean, colour=Treatment)) + 
    geom_errorbar(aes(ymin=mean-se, ymax=mean+se), width=.1) +
    geom_line() +
    geom_point() +
    scale_x_continuous(name="Timepoint") +
    scale_y_continuous(name="Richness") +
    theme_bw() 
    #theme(legend.position="none")
```

Richness dropped off dramtically in Control, f/2+, and DOM+ carboys over the course of all four 10-day incubation experiments. Richness was higher in Control carboys than DOM+ carboys at the midpoint of the August and September experiments. Richness began to rebound in the f/2+ and DOM+ carboys toward the end of the October experiment, perhaps suggesting acclimation. The December experiment results are similar to those from August and September, except that the DOM+ carboys had greater richness than the Control carboys at time 0, suggesting some sort of contamination.

\newpage

# Taxa plots

```{r import_data_for_phyloseq, echo=FALSE, warning=FALSE, include=FALSE, eval=FALSE}
##workflow from here: https://github.com/joey711/phyloseq/issues/336
biom_file = "table_even82638_json.biom" 
map_file = "DOM.map.txt" 
tree_file = "rep_set.tre" 

tree <-read_tree(tree_file)
map <- import_qiime_sample_data(map_file)

mydata <- import_biom(biom_file,tree_file,parseFunction=parse_taxonomy_greengenes)
#warnings(mydata)

#intersect(mydata)
mydata <- merge_phyloseq(mydata,map)

mydata

ntaxa(mydata)
sample_names(mydata)
rank_names(mydata)
sample_variables(mydata)
otu_table(mydata)[1:10, 1:5]
tax_table(mydata)[1:10, 1:5]

get_taxa_unique(mydata, "Phylum")
get_taxa_unique(mydata, "Class")

```

Taxa plots from each of the four experiments suggest that container effects are very strong: similar changes occur in both the control and treatment carboys. There does appear to be a difference in timing. By day 6, community composition in the DOM+ carboys has already shifted dramatically and closely resembles composition at day 10. Changes appear to occur more slowly in the control carboys with composition at day 6 representing a more intermediate state. This proposed lag effect can be seen in the August, September, and December experiments. The data from the October experiment are quite different, as discussed below.


```{r taxa_plots_exp1, echo=FALSE, fig.cap="Changes in abundance (i.e. number of reads) for the twenty most prevalent OTUs over the course of the 10-day incubation period. OTUs are colored by Family.", eval=FALSE}
exp1 <- subset_samples(mydata, Experiment=="1")
TopNOTUs_1 <- names(sort(taxa_sums(exp1), TRUE)[1:20])
exp1a   <- prune_taxa(TopNOTUs_1, exp1)
plot_bar(exp1a, x="Treatment", fill="Family", facet_grid=~Day, title="August")
```

```{r taxa_plot_exp2, echo=FALSE, fig.cap="Changes in abundance (i.e. number of reads) for the twenty most prevalent OTUs over the course of the 10-day incubation period. OTUs are colored by Family.", eval=FALSE}
exp2 <- subset_samples(mydata, Experiment=="2")
TopNOTUs_2 <- names(sort(taxa_sums(exp2), TRUE)[1:20])
exp2a   <- prune_taxa(TopNOTUs_2, exp2)
plot_bar(exp2a, x="Treatment", fill="Family", facet_grid=~Day, title="September")
```

```{r taxa_plot_exp3, echo=FALSE, fig.cap="Changes in abundance (i.e. number of reads) for the twenty most prevalent OTUs over the course of the 10-day incubation period. OTUs are colored by Family.", eval=FALSE}
exp3 <- subset_samples(mydata, Experiment=="3")
TopNOTUs_3 <- names(sort(taxa_sums(exp3), TRUE)[1:20])
exp3a   <- prune_taxa(TopNOTUs_3, exp3)
plot_bar(exp3a, x="Treatment", fill="Family", facet_grid=~Day, title="October")
```

```{r taxa_plot_exp4, echo=FALSE, fig.cap="Changes in abundance (i.e. number of reads) for the twenty most prevalent OTUs over the course of the 10-day incubation period. OTUs are colored by Family.", eval=FALSE}
exp4 <- subset_samples(mydata, Experiment=="4")
TopNOTUs_4 <- names(sort(taxa_sums(exp4), TRUE)[1:20])
exp4a   <- prune_taxa(TopNOTUs_4, exp4)
plot_bar(exp4a, x="Treatment", fill="Family", facet_grid=~Day, title="December")
```

```{r taxa_figure_general_setup, echo=FALSE}

#function to transpose a data table
tdt <- function(inpdt){
  transposed <- t(inpdt[,-1,with=F]);
  colnames(transposed) <- inpdt[[1]];
  transposed <- data.table(transposed, keep.rownames=T);
  setnames(transposed, 1, names(inpdt)[1]);
  return(transposed);
}

# read in the taxa table and order by sample number
taxaL5 <- fread("cdout_DOM_bac/taxa_plots/table_mc64271_sorted_L5.txt", skip=1)

chlor <- grep('Chloroplast', x=taxaL5$"#OTU ID")
idx_chlor <- grep('Chloroplast',x = taxaL5$"#OTU ID", invert=TRUE)
taxaL5_bac <- taxaL5[i=idx_chlor]

taxaL5.transpose <- tdt(taxaL5_bac)

idx_taxa <- order(taxaL5.transpose$'#OTU ID', decreasing=FALSE)
taxa_sorted <- taxaL5.transpose[idx_taxa]

# convert to data frame and transpose again
taxa_sorted.df <- as.data.frame(taxa_sorted)
taxa_t <- t(taxa_sorted.df[,-1])
colnames(taxa_t) <- taxa_sorted.df[,1]

taxa_t.exp1 <- taxa_t[,(1:18)]
taxa_t.exp2 <- taxa_t[,(19:36)]
taxa_t.exp3 <- taxa_t[,(37:64)]
taxa_t.exp4 <- taxa_t[,(65:81)]

meta_sorted.df <- as.data.frame(meta_sorted)
```

##August
At the beginning of the August experiment, Pelagibacteria (alphaproteobacteria), slow-growing oligotrophs, were dominant, but were replaced by gammaproteobacteria (Collwellia) over the course of the experiment. 

A very brief search for Collwellia yielded some interesting results:

1. Collwellia were associated with the BP oil spill and supposedly digested alkanes and aromatics
[1](http://www.ncbi.nlm.nih.gov/pubmed/22616650)
[2](https://books.google.com/books?id=cQANAwAAQBAJ&pg=PA26&lpg=PA26&dq=Collwellia&source=bl&ots=fQ7kEClmmF&sig=Lz1kFGSRtdNgWGzvBZY0fXHBZlE&hl=en&sa=X&ved=0CDUQ6AEwBWoVChMIp879iIH8xgIV1xuSCh1xpQQp#v=onepage&q=Collwellia&f=false)
2. Bidle and Azam conducted experiments in which they allowed diatom detritus to be colonized by natural bacterial assemblages. In one experiment, the pos-incubation assemblage was characterized by y *Collwellia demingiae*. 
[here](http://aslo.net/lo/toc/vol_46/issue_7/1606.pdf)
3. The group contains some known psychrophiles
[here] (http://www.ncbi.nlm.nih.gov/pubmed/16043709)

Flavobacteria also increased over the course of the experiment but did not comprise a large fraction of the community, even at day 10. 


```{r aug_exp1_taxa_plot, echo=FALSE, fig.cap="Relative abundance of most common taxa (family level) during DOM addition experiments."}

# get row means
taxa_t.exp1 <- as.data.frame(taxa_t.exp1)
taxa_t.exp1$average <- apply(taxa_t.exp1, 1, mean)
# apply index to table to sort according to average relative abundance
idx1 <- order(taxa_t.exp1$average, decreasing = TRUE)
taxa_sorted.exp1 <- taxa_t.exp1[idx1, ]

# keep top 11 taxa, everything else is summed under heading "Other"
Other1 <- colSums(taxa_sorted.exp1[(12:572),(1:18)])

#transpose yet again, rename rows, keep only top 11 taxa and add Other column
taxa_sorted_t1 <- t(taxa_sorted.exp1[,-(19)])
#rownames(taxa_sorted_t1) <- meta_sorted$Day[1:18]
taxa_sorted_t_other1 <-taxa_sorted_t1[,(1:11)]
taxa_sorted_t_other1 <- cbind(taxa_sorted_t_other1, Other1)
taxa_sorted_t_other1.df <- as.data.frame(taxa_sorted_t_other1)
taxa_sorted_t_other1.df$Treatment <- meta_sorted.df$Treatment[(1:18)]
taxa_sorted_t_other1.df$Day <-  meta_sorted.df$Day[(1:18)]

m1 <- melt(taxa_sorted_t_other1.df, id=c("Treatment", "Day"), 
      measured=colnames(taxa_sorted_t_other1.df[,(2:11)]))
colnames(m1) <- c("Treatment","Day", "Family", "RelativeAbundance")

#summarize data
tdata1 <- ddply(m1, c("Treatment", "Day", "Family"), summarise,
               N = length(RelativeAbundance),
               mean = mean(RelativeAbundance),
               sd   = sd(RelativeAbundance),
               se   = sd / sqrt(N)
)

tdata1$Day <- as.factor(tdata1$Day)

day_labeller <- function(var, value){
  value <- as.character(value)
  if (var=="Day") { 
    value[value=="0"] <- "Day 0"
    value[value=="6"]   <- "Day 6"
    value[value=="10"]   <- "Day 10"
  }
  return(value)
}


p1 <- ggplot(data = tdata1, aes(x = Treatment, y = mean, fill = Family)) + 
  geom_bar(stat="identity", color="black", size=0.1) +
  scale_fill_brewer(palette="Paired",
                   labels=c("Colwelliaceae",
                            "Pelagibacteraceae", 
                            "Flavobacteriaceae", 
                            "Oceanospirillales-Other",
                            "Rhodobacteraceae", 
                            "Flavobacteriales-NS9",
                            "Alpha-Other", 
                            "Delta-SAR324", 
                            "Nitrospinaceae",
                            "Vibrionaceae",
                            "Oceanospirillaceae",
                            "Other"),
                    name="Taxa",
                   guide = guide_legend(reverse=TRUE)) +
  ylab("Relative abundance") +
  #ggtitle("August") +
  xlab("") +
  theme_bw() +
  theme(text = element_text(size=10),
        axis.text.x = element_text(vjust=1))

p1 + facet_grid(.~Day, labeller=day_labeller)

```

\newpage

##September
Similar trends (declining Pelagibacteria and increasing Collwellia) occurred during the September experiment. Polaribacter (Flavobacteria blocks) was prevalent in both Control and DOM+ carboys on Day 10. Carbon from phytoplankton die-off (as suggested by disappearing chloroplast sequences) could account for some of the changes that occurred in the Control carboys.

```{r sept_exp2_taxa_plot, echo=FALSE, fig.cap="Relative abundance of most common taxa (family level) during DOM addition experiments."}

# get row means
taxa_t.exp2 <- as.data.frame(taxa_t.exp2) 
taxa_t.exp2$average <- apply(taxa_t.exp2, 1, mean) #create a col containing row means
# apply index to table to sort according to average relative abundance
idx2 <- order(taxa_t.exp2$average, decreasing = TRUE)
taxa_sorted.exp2 <- taxa_t.exp2[idx2, ] 

# keep top 11 taxa, everything else is summed under heading "Other"
Other2 <- colSums(taxa_sorted.exp2[(12:572),(1:18)]) #drop col19 (average)

#transpose yet again, rename rows, keep only top 11 taxa and add Other column
taxa_sorted_t2 <- t(taxa_sorted.exp2[,-(19)]) #transpose table w/o average column
taxa_sorted_t_other2 <-taxa_sorted_t2[,(1:11)] #keep only the top 11 taxa
taxa_sorted_t_other2 <- cbind(taxa_sorted_t_other2, Other2) #add Other to table
taxa_sorted_t_other2.df <- 
    as.data.frame(taxa_sorted_t_other2) #convert to df so later stuff works
taxa_sorted_t_other2.df$Treatment <- meta_sorted.df$Treatment[(19:36)] #add info from metadata
taxa_sorted_t_other2.df$Day <-  meta_sorted.df$Day[(19:36)]

# convert data to long format to make barplot
m2 <- melt(taxa_sorted_t_other2.df, id=c("Treatment", "Day"), 
      measured=colnames(taxa_sorted_t_other2.df[,(2:11)]))
colnames(m2) <- c("Treatment","Day", "Family", "RelativeAbundance")

# summarize data
tdata2 <- ddply(m2, c("Treatment", "Day", "Family"), summarise,
               N = length(RelativeAbundance),
               mean = mean(RelativeAbundance),
               sd   = sd(RelativeAbundance),
               se   = sd / sqrt(N)
)

tdata2$Day <- as.factor(tdata1$Day) #change day to actor so x-axis spacing is even

# this function allows you to change facet grid labels
day_labeller <- function(var, value){
  value <- as.character(value)
  if (var=="Day") { 
    value[value=="0"] <- "Day 0"
    value[value=="6"]   <- "Day 6"
    value[value=="10"]   <- "Day 10"
  }
  return(value)
}

p2 <- ggplot(data = tdata2, aes(x = Treatment, y = mean, fill = Family)) + 
  geom_bar(stat="identity", color="black", size=0.1) +
  scale_fill_brewer(palette="Paired",
                   labels=c("Colwelliaceae",
                            "Pelagibacteraceae", 
                            "Flavobacteriaceae",
                            "Oceanospirillales-Other",
                            "Rhodobacteraceae", 
                            "Flavobacteriales-NS9", 
                            "Alpha-Other", 
                            "Nitrospinaceae", 
                            "Delta-SAR324", 
                            "Cryomorphaceae", 
                            "Oceanospirillaceae",
                            "Other"),
                    name="Taxa",
                   guide = guide_legend(reverse=TRUE)) +
  ylab("Relative abundance") +
  xlab("") +
  #ggtitle("September") +
  theme_bw() +
  theme(text = element_text(size=10),
        axis.text.x = element_text(vjust=1))

p2 + facet_grid(.~Day, labeller=day_labeller)

```


##October
The results from the October experiment, which included an f/2+ treatment, are strange. At the beginning of the experiment, Rickettsiales comprises large fractions of all of the communities. The grey blocks are made up of Haptophyceae (chloroplasts) and Oceanospirillales. At day 6, we have no data from the control carboys. The f/2+ and DOM+ carboys are both dominated by flavobacteria. These flavobacter OTUs are classified as Polaribacter. The DOM+ carboys also have a significant block of Alteromonadales. On day 10, the control and f/2+ carboys are very similar, while the DOM+ carboys are dominated to a greater extent by Polaribacter. A few additional samples were collected on day 11, which I find to be suspect. My inclination for this experiment would be to only consider data from day 0 and day 10. Again, carbon from phytoplankton die-off (as suggested by disappearing chloroplast sequences) could account for some of the changes that occurred in the Control carboys.

\newpage

##December
In the December experiment, the day 0 communities had a number of chloroplasts (grey blocks). Both Control and DOM+ carboys acquired Polaribacter (Flavobacteria) over the course of the experiment. Collwellia appeared in the DOM+ carboys but not the Control carboys.

```{r dec_exp4_taxa_plot, echo=FALSE, fig.cap="Relative abundance of most common taxa (family level) during DOM addition experiments."}

# get row means
taxa_t.exp4 <- as.data.frame(taxa_t.exp4)
taxa_t.exp4$average <- apply(taxa_t.exp4, 1, mean)
# apply index to table to sort according to average relative abundance
idx4 <- order(taxa_t.exp4$average, decreasing = TRUE)
taxa_sorted.exp4 <- taxa_t.exp4[idx4, ]

# keep top 11 taxa, everything else is summed under heading "Other"
Other4 <- colSums(taxa_sorted.exp4[(12:572),(1:17)])


#transpose yet again, rename rows, keep only top 11 taxa and add Other column
taxa_sorted_t4 <- t(taxa_sorted.exp4[,(1:17)])
taxa_sorted_t_other4 <-taxa_sorted_t4[,(1:11)]
taxa_sorted_t_other4 <- cbind(taxa_sorted_t_other4, Other4)
taxa_sorted_t_other4.df <- as.data.frame(taxa_sorted_t_other4)
taxa_sorted_t_other4.df$Treatment <- meta_sorted.df$Treatment[(65:81)]
taxa_sorted_t_other4.df$Day <-  meta_sorted.df$Day[(65:81)]

m <- melt(taxa_sorted_t_other4.df, id=c("Treatment", "Day"), 
      measured=colnames(taxa_sorted_t_other4.df[,(1:12)]))
colnames(m) <- c("Treatment","Day", "Family", "RelativeAbundance")

#summarize data
tdata4 <- ddply(m, c("Treatment", "Day", "Family"), summarise,
               N = length(RelativeAbundance),
               mean = mean(RelativeAbundance),
               sd   = sd(RelativeAbundance),
               se   = sd / sqrt(N)
)

tdata4$Day <- as.factor(tdata4$Day)

day_labeller <- function(var, value){
  value <- as.character(value)
  if (var=="Day") { 
    value[value=="0"] <- "Day 0"
    value[value=="6"]   <- "Day 6"
    value[value=="10"]   <- "Day 10"
  }
  return(value)
}


p4 <- ggplot(data = tdata4, aes(x = Treatment, y = mean, fill = Family)) + 
  geom_bar(stat="identity", colour="black", size=0.1) +
  scale_fill_brewer(palette="Paired",
                   labels=c("Flavobacteriaceae","Pelagibacteraceae", 
                            "Rhodobacteraceae","Colwelliaceae",
                              "Oceanospirillales-Other", "Oceanospirillaceae",
                              "Alpha-Other","Altermonadales-Other", 
                              "Piscirickettsiaceae", "Flammeovirgaceae",
                              "Alteromonadaceae","Other"),
                    name="Taxa",
                   guide = guide_legend(reverse=TRUE)) +
  ylab("Relative abundance") +
  #ggtitle("December") +
  xlab("") +
  theme_bw() +
  theme(text = element_text(size=10),
        axis.text.x = element_text(vjust=1))

p4 + facet_grid(.~Day, labeller=day_labeller)

```


#NMDS

```{r NMDS_large_dataset, eval=FALSE}
# for large datasets that crash R
# didn't need for this dataset
# compile the c++ code from within R
Rcpp::sourceCpp('bc_sparse.cpp')

dis <- my_bray_curtis(otutable_bac_samples_t)
m <- monoMDS(dis)
plot(m)
```

```{r NMDS_regular, echo=FALSE, fig.cap="NMDS of samples from four DOM experiments (conducted in August, September, October, and December 2013). Two treatment (Control and DOM+) are shown, one (f/2+ is not). Carboys were samples on days 0, 6, and 10 except in one case when samples were also collected on day 11."}
mds <- metaMDS(otutable_bac_samples_t, dist="bray", k=2, autotransform=F)
coded.mds <- cbind(mds$points, meta_sorted)

coded.mds$Day <- as.factor(coded.mds$Day)
coded.mds$Treatment <- as.factor(coded.mds$Treatment)
coded.mds$Experiment <- as.factor(coded.mds$Experiment)
# p1 <- ggplot(coded.mds, aes(x=MDS1, y=MDS2, color=Treatment, shape=Day)) +
#   geom_point(size=6) +
#   scale_shape_manual(values=c(1, 10, 16, 20)) +
#   scale_color_manual(values=c("#000000", "#D55E00", "red")) +
#   theme_bw()
# p1
# 
# p2 <- ggplot(coded.mds, aes(x=MDS1, y=MDS2, color=Treatment, shape=Experiment)) +
#   geom_point(size=6) +
#   scale_shape_manual(values=c(1, 10, 16, 20)) +
#   scale_color_manual(values=c("#000000", "#D55E00", "red")) +
#   theme_bw()
# p2

plot(mds, display = "sites", choices = c(1, 2),
     type = "n", shrink = FALSE, main="", xlim=c(-1.5, 1.8))


points(coded.mds$MDS1[which(coded.mds$Treatment=="Control" 
                            & coded.mds$Experiment=="1"
                            & coded.mds$Day=="0")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="Control"
                            & coded.mds$Experiment=="1"
                            & coded.mds$Day=="0")], col="black")
points(coded.mds$MDS1[which(coded.mds$Treatment=="Control" 
                            & coded.mds$Experiment=="1"
                            & coded.mds$Day=="6")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="Control"
                            & coded.mds$Experiment=="1"
                            & coded.mds$Day=="6")], col="black", pch=10)
points(coded.mds$MDS1[which(coded.mds$Treatment=="Control" 
                            & coded.mds$Experiment=="1"
                            & coded.mds$Day=="10")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="Control"
                            & coded.mds$Experiment=="1"
                            & coded.mds$Day=="10")], col="black", pch=16)

points(coded.mds$MDS1[which(coded.mds$Treatment=="DOM" 
                            & coded.mds$Experiment=="1"
                            & coded.mds$Day=="0")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="DOM"
                            & coded.mds$Experiment=="1"
                            & coded.mds$Day=="0")], col="black", pch=0)
points(coded.mds$MDS1[which(coded.mds$Treatment=="DOM" 
                            & coded.mds$Experiment=="1"
                            & coded.mds$Day=="6")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="DOM"
                            & coded.mds$Experiment=="1"
                            & coded.mds$Day=="6")], col="black", pch=7)
points(coded.mds$MDS1[which(coded.mds$Treatment=="DOM" 
                            & coded.mds$Experiment=="1"
                            & coded.mds$Day=="10")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="DOM"
                            & coded.mds$Experiment=="1"
                            & coded.mds$Day=="10")], col="black", pch=15)


points(coded.mds$MDS1[which(coded.mds$Treatment=="Control" 
                            & coded.mds$Experiment=="2"
                            & coded.mds$Day=="0")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="Control"
                            & coded.mds$Experiment=="2"
                            & coded.mds$Day=="0")], col="blue")
points(coded.mds$MDS1[which(coded.mds$Treatment=="Control" 
                            & coded.mds$Experiment=="2"
                            & coded.mds$Day=="6")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="Control"
                            & coded.mds$Experiment=="2"
                            & coded.mds$Day=="6")], col="blue", pch=10)
points(coded.mds$MDS1[which(coded.mds$Treatment=="Control" 
                            & coded.mds$Experiment=="2"
                            & coded.mds$Day=="10")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="Control"
                            & coded.mds$Experiment=="2"
                            & coded.mds$Day=="10")], col="blue", pch=16)

points(coded.mds$MDS1[which(coded.mds$Treatment=="DOM" 
                            & coded.mds$Experiment=="2"
                            & coded.mds$Day=="0")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="DOM"
                            & coded.mds$Experiment=="2"
                            & coded.mds$Day=="0")], col="blue", pch=0)
points(coded.mds$MDS1[which(coded.mds$Treatment=="DOM" 
                            & coded.mds$Experiment=="2"
                            & coded.mds$Day=="6")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="DOM"
                            & coded.mds$Experiment=="2"
                            & coded.mds$Day=="6")], col="blue", pch=7)
points(coded.mds$MDS1[which(coded.mds$Treatment=="DOM" 
                            & coded.mds$Experiment=="2"
                            & coded.mds$Day=="10")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="DOM"
                            & coded.mds$Experiment=="2"
                            & coded.mds$Day=="10")], col="blue", pch=15)

points(coded.mds$MDS1[which(coded.mds$Treatment=="Control" 
                            & coded.mds$Experiment=="3"
                            & coded.mds$Day=="0")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="Control"
                            & coded.mds$Experiment=="3"
                            & coded.mds$Day=="0")], col="orange")
points(coded.mds$MDS1[which(coded.mds$Treatment=="Control" 
                            & coded.mds$Experiment=="3"
                            & coded.mds$Day=="6")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="Control"
                            & coded.mds$Experiment=="3"
                            & coded.mds$Day=="6")], col="orange", pch=10)
points(coded.mds$MDS1[which(coded.mds$Treatment=="Control" 
                            & coded.mds$Experiment=="3"
                            & coded.mds$Day=="10")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="Control"
                            & coded.mds$Experiment=="3"
                            & coded.mds$Day=="10")], col="orange", pch=16)

points(coded.mds$MDS1[which(coded.mds$Treatment=="DOM" 
                            & coded.mds$Experiment=="3"
                            & coded.mds$Day=="0")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="DOM"
                            & coded.mds$Experiment=="3"
                            & coded.mds$Day=="0")], col="orange", pch=0)
points(coded.mds$MDS1[which(coded.mds$Treatment=="DOM" 
                            & coded.mds$Experiment=="3"
                            & coded.mds$Day=="6")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="DOM"
                            & coded.mds$Experiment=="3"
                            & coded.mds$Day=="6")], col="orange", pch=7)
points(coded.mds$MDS1[which(coded.mds$Treatment=="DOM" 
                            & coded.mds$Experiment=="3"
                            & coded.mds$Day=="10")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="DOM"
                            & coded.mds$Experiment=="3"
                            & coded.mds$Day=="10")], col="orange", pch=15)

points(coded.mds$MDS1[which(coded.mds$Treatment=="Control" 
                            & coded.mds$Experiment=="4"
                            & coded.mds$Day=="0")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="Control"
                            & coded.mds$Experiment=="4"
                            & coded.mds$Day=="0")], col="red")
points(coded.mds$MDS1[which(coded.mds$Treatment=="Control" 
                            & coded.mds$Experiment=="4"
                            & coded.mds$Day=="6")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="Control"
                            & coded.mds$Experiment=="4"
                            & coded.mds$Day=="6")], col="red", pch=10)
points(coded.mds$MDS1[which(coded.mds$Treatment=="Control" 
                            & coded.mds$Experiment=="4"
                            & coded.mds$Day=="10")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="Control"
                            & coded.mds$Experiment=="4"
                            & coded.mds$Day=="10")], col="red", pch=16)

points(coded.mds$MDS1[which(coded.mds$Treatment=="DOM" 
                            & coded.mds$Experiment=="4"
                            & coded.mds$Day=="0")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="DOM"
                            & coded.mds$Experiment=="4"
                            & coded.mds$Day=="0")], col="red", pch=0)
points(coded.mds$MDS1[which(coded.mds$Treatment=="DOM" 
                            & coded.mds$Experiment=="4"
                            & coded.mds$Day=="6")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="DOM"
                            & coded.mds$Experiment=="4"
                            & coded.mds$Day=="6")], col="red", pch=7)
points(coded.mds$MDS1[which(coded.mds$Treatment=="DOM" 
                            & coded.mds$Experiment=="3"
                            & coded.mds$Day=="10")], 
       coded.mds$MDS2[which(coded.mds$Treatment=="DOM"
                            & coded.mds$Experiment=="3"
                            & coded.mds$Day=="10")], col="red", pch=15)

legend(1.2,0.75, 
       c("Control","DOM+"),
       pch=c(1,0),
       bty="n") 
legend(1.2,0.35, 
       c("Aug","Sept","Oct","Dec"),
       pch=c(16,16,16,16),
       col=c("black", "blue", "orange", "red"),
       bty="n") 
legend(1.2,-0.25, 
       c("Day 0","Day 6", "Day 10", "Day 11"),
       pch=c(0,7,15,15),
       bty="n") 

```

```{r NMDS_multiplot, eval=FALSE, echo=FALSE}
mds <- metaMDS(otutable_bac_samples_t, dist="bray", k=2, autotransform=F)
coded.mds <- cbind(mds$points, meta_sorted)

coded.mds$Day <- as.factor(coded.mds$Day)
coded.mds$Treatment <- as.factor(coded.mds$Treatment)
coded.mds$Experiment <- as.factor(coded.mds$Experiment)

p1 <- ggplot(coded.mds, aes(x=MDS1, y=MDS2, color=Treatment, shape=Day)) +
  geom_point(size=6) +
  scale_shape_manual(values=c(1, 10, 16, 21)) +
  scale_color_manual(values=c("#000000", "#D55E00", "#009E73")) +
  theme_bw()
p1

p2 <- ggplot(coded.mds, aes(x=MDS1, y=MDS2, color=Day)) +
  geom_point(size=6) +
  #scale_shape_manual(values=c(1, 10, 16, 21)) +
  scale_color_manual(values=c("#000000", "#E69F00", "#56B4E9", "#009E73")) +
  theme_bw()
p2

p2 <- ggplot(coded.mds[which(coded.mds$Experiment=="1")], aes(x=MDS1, y=MDS2, color=Treatment)) +
  geom_point(size=6) +
  #scale_shape_manual(values=c(1, 10, 16, 21)) +
  scale_color_manual(values=c("#000000", "#E69F00", "#56B4E9")) +
  theme_bw()
p2

p2 <- ggplot(coded.mds, aes(x=MDS1, y=MDS2, color=Treatment)) +
  geom_point(size=6, alpha=0.5) +
  #scale_shape_manual(values=c(1, 10, 16, 21)) +
  scale_color_manual(values=c("#000000", "#E69F00", "#56B4E9")) +
  theme_bw()
p2

```


##August

The Control and DOM+ carboys are similar to each other on both Day 0 and Day 10. As seen previously in the taxa plots, DOM+ carboys seems to more rapidly approach final Day 10 community composition. Day 6 DOM+ carboys are clustered with Day 10 DOM+ carboys, while the Day 6 control carboys form their own cluster. Note that the replicate carboys do, in fact, cluster together.


```{r, echo=FALSE, eval=FALSE}

coded.mds$Experiment <- as.factor(coded.mds$Experiment)
coded.mds$Day <- as.factor(coded.mds$Day)
coded.mds_1 <- coded.mds[which(coded.mds$Experiment=="1")]
coded.mds_2 <- coded.mds[which(coded.mds$Experiment=="2")]
coded.mds_3 <- coded.mds[which(coded.mds$Experiment=="3")]
coded.mds_4 <- coded.mds[which(coded.mds$Experiment=="4")]

p1 <- ggplot(coded.mds1, aes(x=MDS1, y=MDS2, color=Treatment, shape=Day)) +
  geom_point(size=6) +
  scale_shape_manual(values=c(1, 10, 16)) +
  scale_color_manual(values=c("#000000", "#D55E00")) +
  ggtitle("August") +
  theme_bw()
p1
```

##September

The September NMDS results are similar to the August results. Day 0 and Day 10 carboys all group together regardless of treatment. The Day 6 DOM+ carboys are more similar to the Day 10 cluster, while the Day 6 Control carboys are more similar to the Day 0 cluster.

```{r, echo=FALSE, eval=FALSE}
p2 <- ggplot(coded.mds_2, aes(x=MDS1, y=MDS2, color=Treatment, shape=Day)) +
  geom_point(size=6) +
  scale_shape_manual(values=c(1, 10, 16)) +
  scale_color_manual(values=c("#000000", "#D55E00")) +
  ggtitle("September") +
  theme_bw()
p2
```

##October

Recall the the October experiment sampling scheme had a few discrepancies (no Control of Day 6, additional samples collected on Day 11). In this experiment, the f/2+ carboys clustered with the DOM+ carboys midway through the experiment but clustered with the control carboys at the end of the experiment. The DOM+ carboys changed between Day 0 and Day 6 but did not change after that. 

```{r, echo=FALSE, eval=FALSE}
p3 <- ggplot(coded.mds_3, aes(x=MDS1, y=MDS2, color=Treatment, shape=Day)) +
  geom_point(size=6) +
  scale_shape_manual(values=c(1, 10, 16, 18)) +
  scale_color_manual(values=c("#000000", "#D55E00", "#56B4E9")) +
  ggtitle("October") +
  theme_bw()
p3
```

##December

In this experiment, both the Control carboys and the DOM+ carboys appeared to reach their final states by Day 6.

```{r, echo=FALSE, eval=FALSE}
p4 <- ggplot(coded.mds_4, aes(x=MDS1, y=MDS2, color=Treatment, shape=Day)) +
  geom_point(size=6) +
  scale_shape_manual(values=c(1, 10, 16)) +
  scale_color_manual(values=c("#000000", "#D55E00")) +
  ggtitle("December") +
  theme_bw()
p4

#multiplot(p1,p2,p3,p4, cols=2)
```

##Add ellipses (not working right now)

```{r NMDS_w_ellipses, echo=FALSE, , eval=FALSE, fig.cap="NMDS without PAL0910. Ellipses represent 95% confidence intervals."}
#Get MDS stats
otutable_bac_samples_t1 <- otutable_bac_samples_t[(1:18),]
meta_sorted1 <- meta_sorted[(1:18),]
mds1 <- metaMDS(otutable_bac_samples_t1, dist="horn", k=2, autotransform=F)
coded.mds1 <- cbind(mds$points, meta_sorted)

# coded.mds.exp1 <- coded.mds[which(coded.mds$Experiment=="1")]
# coded.mds.exp1 <- as.data.frame(coded.mds.exp1)
# mds.exp1 <- coded.mds.exp1[,(1:3)]

#Make a new data frame, and put Season and Month there
# to be useful for coloring, and shape of points
NMDS.Exp1=data.frame(x=mds1$point[,1],
                y=mds1$point[,2],
                Treatment=as.factor(meta_sorted1$Treatment),
                Day=as.factor(meta_sorted1$Day))

#Get spread of points based on season
plot.new()
ord <- ordiellipse(mds1, meta_sorted1$Day,
                 display = "sites", kind ="sd", conf = 0.95, label = T)


#Reference: http://stackoverflow.com/questions/13794419/plotting-ordiellipse-function-from-vegan-package-onto-nmds-plot-created-in-ggplo
#Data frame df_ell contains values to show ellipses. 
#It is calculated with function veganCovEllipse which is hidden in vegan package. 
#This function is applied to each level of NMDS (group) 
#and it uses also function cov.wt to calculate covariance matrix.

veganCovEllipse<-function (cov, center = c(0, 0), scale = 1, npoints = 100) 
{
  theta <- (0:npoints) * 2 * pi/npoints
  Circle <- cbind(cos(theta), sin(theta))
  t(center + scale * t(Circle %*% chol(cov)))
}

#Generate ellipse points
df_ell <- data.frame()
for(g in levels(NMDS.Exp1$Day)){
  if(g!="" && (g %in% names(ord))){
    
    df_ell <- 
      rbind(df_ell, cbind(as.data.frame(with(NMDS.Exp1[NMDS.Exp1$Day==g,],
      veganCovEllipse(ord[[g]]$cov,ord[[g]]$center,ord[[g]]$scale)))
      ,Season=g))
  }
}

head(df_ell)

#Generate mean values from NMDS plot grouped on Countries
NMDS.Exp1.mean=aggregate(NMDS.Exp1[,1:2],list(group=NMDS.Exp1$Day),mean)
NMDS.Exp1.mean

#plot

ggplot(data=NMDS.Exp1,aes(x,y,colour="Day")) +
  annotate("text",x=NMDS.Exp1.mean$x,y=NMDS.Exp1.mean$y,label=NMDS.Exp1.mean$group,size=4) +
  geom_path(data=df_ell, aes(x=NMDS1, y=NMDS2), size=1, linetype=2) +
  geom_point(aes(shape=Treatment)) +
  theme_bw() 
```
